---
title: "RHICCA Data Merge"
author: "Olivia Swarthout"
date: "July 31, 2020"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = '')
rm(list = ls())
library(tidyverse)
library(dplyr)
library(plyr)
library(readr)
library(tibble)
library(kableExtra)

mydir <- "RHICCA/FOLLOWUP_6MONTHS"

# read in .txt file of uncertain IDs as list
uncertain <- scan("C:\\Users\\Olivia Swarthout\\Desktop\\uncertain_6months.txt", what = character(), sep = ",")

```

This is a script to merge Luminex data tables from the RHICCA Project into a single table for each timepoint at which samples were collected. The code accounts for the potential of small variations in the naming of spreadsheets and their columns, but does need the input sheets to have all columns in the same order, and to include "plate x" (x being the relevant plate number) somewhere in their file paths. 

All data should be stored in CSV form or else it won't be detected. The variable mydir above should point to the folder for the single timpoint which is being merged, as the code reads in all .csv files within mydir. However, if the aim is to combine data from multiple timepoints within one spreadsheet, the code will accomplish this while including a column with the timepoint for each sample that is read in so as to distinguish between timepoints.

```{r fileprep, include=FALSE}

# read in all .csv file paths from directory
plates = list.files(path = mydir, pattern = "*.csv", full.names=TRUE, recursive = TRUE, 
                    include.dirs = TRUE ) 


#create an empty data frame to hold all merged data
merged <- data.frame("SampleID"=character(), "PlateID" = character(), "CD163" = character(),
                     "ICAM1"=character(), "ILb"=character(), "IL6"=character())

# get rid of "questionable results"
plates <- Filter(function(x) !any(grepl("Questionable", x)), plates)

```



### Plates  
Below is the list of plates read in by the code, excluding those labelled as "questionable results" and those plates for which a repeat was later run. Issues with this list (plates missing or included when they should not be) may arise from inconsistencies in the naming of files and folders.

```{r plates, echo=FALSE}
print(plates)

```

Reading and merging the data now

```{r merge}

 for (x in plates[1:22]){
  current <- read.csv(x)
  
  #converts entire table to type character  
  current[] <- lapply(current, as.character)
  
  #some tables had slight discrepancies in column names so this will account for that by renaming 
  names(current)[1:6] <- c("SampleID", "PlateID", "CD163", "ICAM1", "ILb", "IL6")

  #a regex to extract the subdirectory name from the full path
  platenum <- str_extract(x, regex("PLATE \\-*\\d+\\.*\\d*", ignore_case = TRUE))
 
  #take just plate number from subdirectory name, store in plate ID column
  platenum <- gsub(platenum, pattern = "PLATE ", replacement = "", ignore.case = TRUE)
  
  #get timepoint by finding folder directly inside rhicca folder
 
  time <- str_extract(x, "RHICCA\\/(\\d|\\w|_)*" )
  time <- gsub(time, pattern = "RHICCA/", replacement = "")
  
  current$PlateID <- paste(time, platenum, sep = "_")
  
  #add new column containing timepoint
  current$timepoint <- time
  
  merged <- bind_rows(merged, current)
 }
```
 
### QC Rows
Removing rows with QC IDs, which exist for quality control purposes

```{r QC}
s <- which(merged$SampleID == "QC 1"| merged$SampleID == "QC 2" | merged$SampleID == "QC1"|merged$SampleID == "QC2")
merged <- merged[-s,]
```

```{r uncertain}
#flag IDs that were in red text (i.e. validity is uncertain)
#add a metadata column if one does not already exist
if(ncol(merged)<8){
  merged$metadata <- NA
}else{
  names(merged)[8] <- "metadata"
}

#label missing or uncertain IDs (using list of IDs read in from .txt)
p <- which(!is.na(match(merged$SampleID, uncertain)) | merged$SampleID == "")
merged$metadata[p] <- "uncertain/missing ID"

```

### Samples that fall on detection limit

Before it looks for the samples that fall above or below the detection limit, this program looks for samples that fall exactly on the limit, and prints them out if any exist.  

```{r onlimit}
t <- grep("(?<!<\\ )2.44140625", merged$ILb, perl = TRUE, value = FALSE)
if(length(t)==0){
  print("No ILb values lie on the lower detection limit")
}else{
  print("The following samples have ILb value at detection limit:")
  merged[t,]
}
u <- grep("(?<!<\\ )2.44140625", merged$IL6, perl = TRUE, value = FALSE)
if(length(u)==0){
  print("No IL6 values lie on the lower detection limit")
}else{
  print("The following samples have IL6 value at detection limit:")
  merged[u,]
}
x <- grep("(?<!>\\ )1000", merged$CD163, perl = TRUE, value = FALSE)
if(length(x)==0){
  print("No CD163 values lie on the upper detection limit")
}else{
  print("The following samples have CD163 value at upper detection limit:")
  merged[x,]
}
y <- grep("(?<!>\\ )250000", merged$ICAM1, perl = TRUE, value = FALSE)
if(length(y)==0){
  print("No ICAM1 values lie on the upper detection limit")
}else{
  print("The following samples have ICAM1 value at upper detection limit:")
  merged[y,]
}

```


### Samples above and below detection limits
 The number of samples out of the total that do not fall within that range of detection are printed below. 
 For each of the four biomarkers, a column is created in the spreadsheet
 
 Their values are then set to that

```{r imputelimits}
#find indices which contain a < or >
b <-  grepl(">",merged$CD163)
#create new column which indicates if above or below limit
merged$CD163Limit[b] <- "above"
#print number that are censored
cat(sum(b, na.rm = TRUE), "samples out of", nrow(merged), "have CD163 levels above detection limit")
#impute censored values to limit
merged$CD163[b] <- 1000

c <- grepl(">",merged$ICAM1)
merged$ICAM1Limit[c] <- "above"
cat(sum(c, na.rm = TRUE), "samples out of", nrow(merged), "have ICAM1 levels above detection limit")
merged$ICAM1[c] <- 250000

d <- grepl("<",merged$ILb)
merged$ILbLimit[d] <- "below"
cat(sum(d, na.rm = TRUE), "samples out of", nrow(merged), "have ILb levels below detection limit")
merged$ILb[d] <- 2.44140625

e <-  grepl("<",merged$IL6)
merged$IL6Limit[e] <- "below" 
cat(sum(e, na.rm = TRUE), "samples out of", nrow(merged), "have IL6 levels below detection limit")
merged$IL6[e] <- 2.44140625


#when we converted the table to type character, this converted NA to "N/A" -- change this back so we can convert back to numeric
merged[merged == "N/A"] <- NA 

```


Converting biomarker columns to type numeric

```{r asnumeric, warning=FALSE}
#converts all biomarker values to type numeric
merged$CD163 <- as.numeric(merged$CD163, na.rm = TRUE)
merged$ICAM1 <- as.numeric(merged$ICAM1, na.rm = TRUE)
merged$IL6 <- as.numeric(merged$IL6, na.rm=TRUE)
merged$ILb <- as.numeric(merged$ILb, na.rm=TRUE)

```

### Duplicated Patient IDs

All duplicated IDs and their frequencies printed below. 

Number of missing IDs printed separately

If a duplicated ID is from a plate in a "REPEATS" folder, the original is removed from the data frame.

```{r findduplicates}
#find number of missing sample IDs
indices <- which(merged$SampleID == "")
cat(length(indices), "samples have missing ID: indices", paste(as.character(indices), collapse = ", "))


#create table with non-missing IDs that are repeated
merged$SampleID[merged$SampleID == ""] <- NA

#create frequency table for SampleID columns, subset to only frequencies greater than 1
n_occur <- data.frame(table(merged$SampleID))
duplicates <- n_occur[n_occur$Freq > 1,]
duplicates$Var1 <- factor(duplicates$Var1)
  kable(duplicates, caption = "Sample IDs With More Than One Occurrence") %>%
  kable_styling()
```

Removing duplicate IDs

```{r removeduplicates}

#find plate IDs of repeat plates, reconstruct timepoint identifier
repeats <- grep("REPEATS", plates, value = TRUE)
repeatplates  <- str_extract(repeats, regex("PLATE \\-*\\d+\\.*\\d*", ignore_case = TRUE))
repeatplates <- paste(time, repeatplates, sep = "_")
repeatplates <- gsub(repeatplates, pattern = "PLATE ", replacement = "")

#iterate through duplicates
for (item in duplicates$Var1){
  #for each duplicated ID, find all matching rows
  current <- which(merged$SampleID == item)
  #set indicator boolean, iterate through all matches of the id
  isrep <- FALSE
    for (ind in current){
      if (merged$PlateID[ind] %in% repeatplates){
        #indicate true if one of the matches is from a repeats plate
        isrep <- TRUE
      }
    }
    #if one is from the repeats plate, throw out the other ones that are not repeats
    if(isrep == TRUE){
    for (ind in current) {
      if((merged$PlateID[ind] %in% repeatplates) == FALSE)
        cat("sample with ID", merged$SampleID[ind], "found on plate", merged$PlateID[ind], "removed")
        
        merged <- merged[-ind,]
    }
}
}

```

Some incorrect values exist in the data due to machine or human error. These are corrected by the script; however for future timepoints this code will need to be updated 

```{r}
if("BASELINE" %in% merged$timepoint){
  merged$ICAM1[merged$SampleID == "RE10365"] <- 239273.0207
  merged$ILb[merged$SampleID == "RE1239K"] <- NA
  #using value instead of ID as this one has missing ID
  merged$CD163[merged$CD163 == 105099.230098485] <- 599.9892256
}

if("FOLLOWUP_6MONTHS" %in% merged$timepoint){
  merged$CD163[merged$SampleID == "RE19519"] <- 1764.44671
  merged$CD163[merged$SampleID == "RE19535"] <- 3257.210752
  merged$ICAM1[merged$SampleID == "RE1972X"] <- 1807796.497
  merged$ICAM1[merged$SampleID == "RE1907H"] <- NA
}



```


The beginning of the table of merged data can be found below. If the program runs correctly, 12 columns should be shown, representing:

* Sample ID
* Plate ID
* The four biomarkers
* Timepoint
* Metadata (e.g. missing/uncertain)
* An indicator variable for each biomarker stating "above" if the value is above the limit of detection and "below" if below (NA if neither)


```{r}
kable(head(merged)) %>%
  kable_styling()
```
Saving spreadsheet of all merged data as .csv in timepoint folder (can change where it saves)
```{r}
filepath <- paste(time, "_MERGED.csv", sep="")
write.csv(merged, filepath, row.names = FALSE)

cat(filepath, "with", nrow(merged), "lines of data created in working directory")

```

